Physcraper: a python package for continual update of evolutionary estimates using the Open Tree of Life

### 2020-07-10 2-body.Rmd
intro edit first 3 paragraphs

Phylogenetic estimates of evolutionary relationships capture the shared history of living organisms, and provide context for biological observations.
Public biological databases provide an amazing resource for evolutionary estimation, but there is a lot of sequence data available that has never been incorporated into any phylogenetic estimates.

GenBank contains over 217 million DNA sequences [(release number 238; June 15, 2020)](https://ftp.ncbi.nih.gov/genbank/gbrel.txt).
These sequences are estimated to have the potential to resolve evolutionary relationships of most of the 241 000 distinct taxa in GenBank [about `r round(236023/240708, digits=4)*100`% of taxa in the NCBI taxonomy; @sanderson2008phylota].
This in turns covers about 10% of extant described biodiversity [taking a conservative estimate of extant diversity; @scott2011ncbi; @federhen2003taxonomy].

Yet, publicly available estimates of phylogenetic relationships still cover only about 100 000 taxa [@piel2009treebase; @hinchliff2015synthesis; @opentreeoflife2019synth].
These data comprise less than one third of the taxonomic diversity for which sequence data was available a decade ago!

Some of this discrepancy between data availability and phylogenetic estimates is because many phylogenies which are generated and published are not publicly shared in an accesible way [@drew2013lost; @magee2014dawn; @mctavish2018bioessay]. However, there is also a lag between the amount of new DNA data generated and the analysis of these data in a phylogenetic context. Connecting new sequence data with existing estimates of evolutionary relationships is a challenge.

@@

Automated sequence gathering pipelines have been an important incorporation to the field of phylogenetics in many ways, particularly because they represent a clear step towards reproducibility in the field. In contrast, most published phylogenies to date have been inferred using alignments that have been assembled and curated "by hand" [@morrison2009would].

Automated approaches often suffer from errors in homology inference that could rapidly caught by human oversight. CITATION

>>>>

Phylogenetic estimates of evolutionary relationships capture the shared history of living organisms, and provide key context for all our biological observations.
Public biological databases provide an amazing resource for evolutionary estimation, but a large portion of sequence data available has never been incorporated into any phylogenetic estimate.

GenBank, the USA National Center for Biodiversity Information (NCBI) molecular database, release number 159 (April 15, 2007) hosts 72 million DNA sequences that are estimated to have the potential to resolve evolutionary relationships of most of its 241 000 distinct taxa [about `r round(236023/240708, digits=4)*100`% of taxa in the NCBI taxonomy release 159; @sanderson2008phylota].
Yet, current public estimates of phylogenetic relationships are available for only about 100 000 taxa [@piel2009treebase; @hinchliff2015synthesis; @opentreeoflife2019synth], representing less than half of the taxonomic diversity with phylogenetically informative sequence data available in GenBank more than a decade ago.

Some of the discrepancy between data availability and phylogenetic estimates can be explained by the many phylogenies that are generated and published and not shared publicly in an accesible way [@drew2013lost; @magee2014dawn; @mctavish2018bioessay]. However, there is also a lag between the amount of new DNA data generated and the analysis of these data in a phylogenetic context. Connecting new sequence data with existing estimates of evolutionary relationships is a challenge.

@@

### 2020-07-10 2.1-intro-ideas.Rmd
intro edit first 3 paragraphs

>>>>

over 217 million DNA sequences [(release number 238; June 15, 2020)](https://ftp.ncbi.nih.gov/genbank/gbrel.txt)

This in turns covers about 10% of extant described biodiversity [taking a conservative estimate of extant diversity; @scott2011ncbi; @federhen2003taxonomy].

Automated sequence gathering pipelines have been an important incorporation to the field of phylogenetics in many ways, particularly because they represent a clear step towards reproducibility in the field. In contrast, most published phylogenies to date have been inferred using alignments that have been assembled and curated "by hand" [@morrison2009would].

Automated approaches often suffer from errors in homology inference that could rapidly caught by human oversight. CITATION

### 2020-07-10 docs/submission-mee-2020-07-10.tex
mk new affiliation
intro edit first 3 paragraphs

email: \href{mailto:martha.kandziora@mailbox.org}{\nolinkurl{martha.kandziora@mailbox.org}}

@@

\textbf{Word count}: 3768

@@

\textbf{Keywords}: cross-connectivity, gene tree, open science, open tree of life, phylogeny, public database, python, reproducibility, taxonomy, updated alignment

@@

Phylogenetic estimates of evolutionary relationships capture the shared history of living organisms, and provide context for biological observations.
Public biological databases provide an amazing resource for evolutionary estimation, but there is a lot of sequence data available that has never been incorporated into any phylogenetic estimates.

GenBank contains over 217 million DNA sequences \href{https://ftp.ncbi.nih.gov/genbank/gbrel.txt}{(release number 238; June 15, 2020)}.
These sequences are estimated to have the potential to resolve evolutionary relationships of most of the 241 000 distinct taxa in GenBank (about 98.05\% of taxa in the NCBI taxonomy; Sanderson \emph{et al.} \protect\hyperlink{ref-sanderson2008phylota}{2008}).
This in turns covers about 10\% of extant described biodiversity (taking a conservative estimate of extant diversity; Scott \protect\hyperlink{ref-scott2011ncbi}{2011}; Federhen \protect\hyperlink{ref-federhen2003taxonomy}{2003}).

Yet, publicly available estimates of phylogenetic relationships still cover only about 100 000 taxa (Piel \emph{et al.} \protect\hyperlink{ref-piel2009treebase}{2009}; Hinchliff \emph{et al.} \protect\hyperlink{ref-hinchliff2015synthesis}{2015}; OpenTreeOfLife \emph{et al.} \protect\hyperlink{ref-opentreeoflife2019synth}{2019}).
These data comprise less than one third of the taxonomic diversity for which sequence data was available a decade ago!

Some of this discrepancy between data availability and phylogenetic estimates is because many phylogenies which are generated and published are not publicly shared in an accesible way (Drew \emph{et al.} \protect\hyperlink{ref-drew2013lost}{2013}; Magee \emph{et al.} \protect\hyperlink{ref-magee2014dawn}{2014}; McTavish \emph{et al.} \protect\hyperlink{ref-mctavish2018bioessay}{2017}). However, there is also a lag between the amount of new DNA data generated and the analysis of these data in a phylogenetic context. Connecting new sequence data with existing estimates of evolutionary relationships is a challenge.

@@

Automated sequence gathering pipelines have been an important incorporation to the field of phylogenetics in many ways, particularly because they represent a clear step towards reproducibility in the field. In contrast, most published phylogenies to date have been inferred using alignments that have been assembled and curated ``by hand'' (Morrison \protect\hyperlink{ref-morrison2009would}{2009}).

Automated approaches often suffer from errors in homology inference that could rapidly caught by human oversight. CITATION

@@

\leavevmode\hypertarget{ref-federhen2003taxonomy}{}%
Federhen, S. (2003). The taxonomy project. \emph{The NCBI Handbook}.

@@

\leavevmode\hypertarget{ref-morrison2009would}{}%
Morrison, D.A. (2009). Why would phylogeneticists ignore computerized sequence alignment? \emph{Systematic biology}, \textbf{58}, 150--158.

@@

\leavevmode\hypertarget{ref-scott2011ncbi}{}%
Scott, F. (2011). The ncbi taxonomy database. \emph{Nucleic Acids Research}, \textbf{40}, D136--D14.

>>>>

Department of Botany, Faculty of Science, Charles University, Prague, Czech Republic

email: \href{mailto:kandziom@natur.cuni.cz}{\nolinkurl{kandziom@natur.cuni.cz}}

@@

\textbf{Word count}: 3702

@@

\textbf{Keywords}: gene tree, interoperability, open science, open tree of life, phylogeny, public database, python, reproducibility, taxonomy, updated alignment

@@

Phylogenetic estimates of evolutionary relationships capture the shared history of living organisms, and provide key context for all our biological observations.
Public biological databases provide an amazing resource for evolutionary estimation, but a large portion of sequence data available has never been incorporated into any phylogenetic estimate DO WE NEED CITATIONS FOR THIS?.

GenBank, the National Center for Biodiversity Information (NCBI) molecular database, release number 159 (from April 15, 2007) hosts 72 million DNA sequences that are estimated to have the potential to resolve evolutionary relationships of most of its 241 000 distinct taxa (about 98.05\% of taxa in the NCBI taxonomy release 159; Sanderson \emph{et al.} \protect\hyperlink{ref-sanderson2008phylota}{2008}).

Yet, current public estimates of phylogenetic relationships are available for only about 100 000 taxa (Piel \emph{et al.} \protect\hyperlink{ref-piel2009treebase}{2009}; Hinchliff \emph{et al.} \protect\hyperlink{ref-hinchliff2015synthesis}{2015}; OpenTreeOfLife \emph{et al.} \protect\hyperlink{ref-opentreeoflife2019synth}{2019}), representing less than half of the taxonomic diversity with phylogeentically informative sequence data available in GenBank more than a decade ago.

Some of the discrepancy between data availability and phylogenetic estimates can be explained by the many phylogenies that are generated and published and not shared publicly in an accesible way (Drew \emph{et al.} \protect\hyperlink{ref-drew2013lost}{2013}; Magee \emph{et al.} \protect\hyperlink{ref-magee2014dawn}{2014}; McTavish \emph{et al.} \protect\hyperlink{ref-mctavish2018bioessay}{2017}). However, there is also a lag between the amount of new DNA data generated and the analysis of these data in a phylogenetic context. Connecting new sequence data with existing estimates of evolutionary relationships is a challenge.

@@

@@

@@

@@
